Probabilistic Forecasting of the Disturbance Storm Time Index: An Autoregressive Gaussian Process approach

### 2017-05-29 revised-draft.tex
Added new references

%Talk about NARMAX Dst
Another important empirical model used to predict $Dst$ is the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR). The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective function. It must be noted that the NARMAX methodology is not limited to polynomial functions, rather any set of basis function expansions can be used with it, such as but not limited to radial basis functions, wavelets etc \citep{doi:10.1080/00207720600903011}, \citep{JGRA:JGRA17327}. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

@@

@@

Although solar wind magnetospheric coupling functions have been shown to have correlation with geomagnetic indices, they are expressible in terms of $V$ and $B_z$, and hence we do not need to include them as explicit inputs to the model. \emph{Gaussian Process} models derive their strength from automatic feature construction achieved by the covariance functions (interested readers may refer to chapters 6 and 7  in \cite{Rasmussen:2005:GPM:1162254}). Therefore as long as coupling functions can be approximated in the \emph{eigenspace} of the covariance function we need not make them explicit in the input features.

@@

@@

Mean functions in GPR models encode trends in the data, they are the baseline predictions the model falls back to in case the training and test data have little correlation as predicted by the kernel function. If there is no prior knowledge about the function to be approximated, \citet{Rasmussen:2005:GPM:1162254} state that it is perfectly reasonable to choose $m(\mathbf{x} = 0)$ as the mean function, as long as the target values are normalized. In the case of the $Dst$ time series, it is known that the so called \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$ performs quite well in the context of OSA prediction. We therefore choose the \emph{persistence model} as the mean function in our OSA Dst models.

@@

Our results can be summarized as follows.
\begin{enumerate}
      \item \emph{Persistence} model plays an important role in the model building and evaluation process in the context of \emph{one step ahead} prediction of the $Dst$ index. It is clear that the persistence behavior in the $Dst$ values is very strong i.e. the trivial predictive model $\hat{Dst}(t) = Dst(t-1)$ gives excellent performance according to the metrics chosen.

>>>>

%Talk about NARMAX Dst
Another important empirical model used to predict $Dst$ is the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR). The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective function. It must be noted that the NARMAX methodology is not limited to polynomial functions, rather any set of basis function expansions can be used with it, such as radial basis functions, wavelets etc \citep{doi:10.1080/00207720600903011}, \citep{JGRA:JGRA17327}. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

@@

%Local neurofuzzy modelling and singular spectrum analysis.
Apart from the NARMAX and neural network approaches, fuzzy methods have also been applied for $Dst$ prediction, \citet{SWE:SWE146} and \citet{Sharifi2006} outline the application of \emph{Local Neurofuzzy} models for one hour and two hour predictions of $Dst$ respectively. Local neuro-fuzzy models reduce the input space into a number of regions each with its own expert predictor. The combined model predicts $Dst$ for a new point as a linear combinations of the predictions from each expert weighted by a fuzzy score signifying the importance of each model for the provided input. For improving predictive performance of two hour $Dst$ forecasts the authors use \emph{singular spectrum analysis} SSA \citet{loskutov2001testing}, \citet{loskutov2001study}.

@@

Although solar wind magnetospheric coupling functions have been shown to have correlation with geomagnetic indices, they are expressed in terms of $V$ and $B_z$, and hence we do not include them as explicit inputs to the model. \emph{Gaussian Process} models derive their strength from automatic feature construction achieved by the covariance functions (interested readers may refer to chapters 6 and 7  in \cite{Rasmussen:2005:GPM:1162254}). As long as coupling functions can be approximated in the \emph{eigenspace} of the covariance function we need not make them explicit in the input features.

@@

It is an important question as to how unaccounted inputs such as solar wind dynamic pressure $P$ and clock angle $\theta$ affect the structure of the GP-ARX model. From a model selection perspective, these unaccounted inputs should lead to higher values of the noise covariance. In the specific case of solar wind dynamic pressure, it is calculated as a product of the plasma density and the solar wind speed making it highly correlated with the solar wind speed, as a result the GP-ARX model can infer a large portion of the information content from the solar wind speed itself. With respect to the clock angle, it must be noted that coupling functions such as Akasofu $\epsilon$ generally contain powers of $sin \theta$ bounding the effect of clock angle to an absolute magnitude of $1$, hence we do not expect these unaccounted inputs to greatly improve the predictive capabilities of the GP-ARX model.

@@

Mean functions in GPR models encode trends in the data, they are the baseline predictions the model falls back to in case the training and test data have little correlation as predicted by the kernel function. If there is no prior knowledge about the function to be approximated, \citet{Rasmussen:2005:GPM:1162254} state that it is perfectly reasonable to choose $m(\mathbf{x} = 0)$ as the mean function, as long as the target values are normalized. In the case of the $Dst$ time series, it is known that the so called \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$ has high correlation with \emph{OSA} $Dst$. Due to its simplicity, we choose the \emph{persistence model} as the prior mean function in our OSA Dst models.

The \emph{persistence model} can be described as \emph{Markovian} prediction mechanism, when it is chosen as the prior mean of the GP-AR and GP-ARX model, it is indeed the case that the prior probability distribution of $Dst(t)$ is Gaussian with a strong \emph{Markovian} behavior $P(Dst(t)|\mathbf{x}_t) \sim \mathcal{N}(Dst(t-1), \sqrt{K_{osa}(\mathbf{x}_t, \mathbf{x}_t)})$, but the posterior predictive distribution of $Dst(t)$ conditional on the model training data (given in equation \ref{eq:posteriormean}) is non-\emph{Markovian} due its dependence on the term denoted by $\mathbf{K}_{*}$ which contains kernel values computed between the test data and training data features. Thus the GP-AR and GP-ARX models when used conditional on training data are non-\emph{Markovian} predictive models.

@@

Our results can be summarized as follows.
\begin{enumerate}
      \item \emph{Persistence} model plays an important role in the model building and evaluation process in the context of \emph{one step ahead} prediction of the $Dst$ index. Although it is not a robust predictor for the onset of intense geomagnetic storms, the \emph{persistence model} performs well on classical error metrics such as \emph{root mean square error} and such. From the considerations above, it is quite evident that classical performance metrics are not adequate for model evaluation, nevertheless in space weather literature, metrics such as \emph{RMSE} are very commonly used to compare predictive performance of models. Although not the research focus of this study, we note that there exists a need for the formulation of more informative performance metrics for measurement of predictive performance of geomagnetic predictive models.
